---
title: "Background Module Requirements"
output: html_notebook
---

# Compitencies
This document list competencies taught in the Background module


## Assignment 1
* Use R to read a data set from the web
* Use dplyr to create new variables, select variables, and observations.
* Use dplyr to summarize data.
* Use ggplot to plot data
* Use the functional programming of apply, lapply, sapply, tapply and dplyr::purrr.

## Maximum Likelihood.
* Use the principle of maximum likelihood to compare data to hypothesis. 
* Estimate population parameters for continuous and discrete distribution. 

## Flip of a coin

Outcome   | Probability
----------|------------
Heads     | 0.5
Tails     | 0.5


Data | Encoding
-----|--------
Heads| 1

```{r}
likelyhood <- function(heads){
  heads * 0.5 
}
likelyhood(1)
```
Data | Encoding | Prob
-----|----------|-----
Heads| 1        |  0.5
Tails| 0        |  0.5
```{r}
# heads
p1 <- 0.5
# tails
p2 <- 0.5
# Likelihood
p1 * p2
```
Try to find a larger likelihood. This is impossible
Data | Encoding | Prob
-----|----------|-----
Heads| 1        |  0.4
Tails| 0        |  0.6
```{r}
# heads
p1 <- 0.4
# tails
p2 <- 0.6
# Likelihood
p1 * p2
```
Lets look over a grid of values to find a better probability for heads.
First set up a grid of probabilities
```{r}
head_prob <-seq(0,1,0.01)
tail_prob <- 1 - head_prob
tail_prob
```
Now check the likelihood at one point on the grid.  
```{r}
#count the number of heads and the number of tails: one head , one tail
x <-c(1,1)
x[1] * head_prob[1]
x[2] * tail_prob[1]
x[1] * head_prob[1] * x[2] * tail_prob[1]
```
Create a function to do an elementary calculation, so more values can be tested.
Notice the individual probabilities are multiplied together.
```{r}
like <- function(x,p_selection){
   (head_prob[p_selection])^x[1] * tail_prob[p_selection]^x[2] 
}
like(c(1,1),1)
like(c(1,1),2)
like(c(1,1),3)
```
Use functional programming to make testing automatic
```{r}
lapply(1:100,like,x=c(1,1))
```

Consider a different data set with two head but no tails.
Try to find a larger likelihood. This is impossible

Data | Count    | Prob  | Encoding
-----|----------|-------|---------
Heads| 2        |  ?    | 2 
Tails| 0        |  ?    | 0

Lets guess that the probability of tails is close to zero.  Notice that tails never occurs
```{r}
# Encode the data with counts of heads and counts of tails
x <- c( 2,0)
lapply(1:100,like,x=x)
```


Try a different data set.

Data | Count    | Prob  | Encoding
-----|----------|-------|---------
Heads| 10        |  ?    | 10
Tails| 20        |  ?    | 20

There are twice as many tails as heads, so guess that the probability to tails is twice the probability of heads.
```{r}
x <- c( 10,20)
lapply(1:100,like,x=x)
```

Here is the maximum likelihood probabilities for heads and tails.
```{r}
head_prob[34]
tail_prob[34]
```
Exercise:
Replicate this for a three sided dice, then a six sided dice, then for the normal distribution
See ?pnorm.





