---
title: "Background Module Requirements"
output: html_notebook
---

# Compitencies
Here is a review of competencies from the previous module as well as
what is to be understood in this current module.

## Assignment 1
* Use R to read a data set from the web
* Use dplyr to create new variables, select variables, and observations.
* Use dplyr to summarize data.
* Use ggplot to plot data
* Use the functional programming of apply, lapply, sapply, tapply and dplyr::purrr.

## Maximum Likelihood.
* Use the principle of maximum likelihood to compare data to hypothesis. 
* Estimate population parameters for continuous and discrete distribution. 

Likelihood is calculated from the data and the likelihood function.
The parameters of the likelihood function can come from 
a hypothesis about the data.  The hypothesis can quantify
prior expectation, a guess or even what the researcher is trying to 
disprove.  

In the case of matching a child with their parents we associate a 
high probability to children being of the same nationality or 
ethnic background as the parents.  There is a smaller chance the 
child is adopted.  But if we have to make a guess, we go with the 
explaination that has the highest change of occuring.

For our calculations, will will start of with a very simple example and 
calculate the likelihood.

## Flip of a coin

Outcome   | Probability
----------|------------
Heads     | 0.5
Tails     | 0.5

Assume there is one toss and a head is thrown.
```{r}
1 * 0.5
```
Now consider two events.  In the first event a head is tossed,
In the second event a tails is tossed.  The probabilities for the 
two events are multiplied to get the likelihood of the data given the
proposed probabilities given in the table below.

Data | Encoding | Prob
-----|----------|-----
Heads| 1        |  0.5
Tails| 0        |  0.5


```{r}
# heads
p1 <- 0.5
# tails
p2 <- 0.5
# Likelihood
p1 * p2
```
Try to find a larger likelihood by using different probabilities for
the event of head and the event of tails.  The probabilities have to 
add up to one. The pobabilities on the table below will be tried.

Data | Encoding | Prob
-----|----------|-----
Heads| 1        |  0.4
Tails| 0        |  0.6
```{r}
# heads
p1 <- 0.4
# tails
p2 <- 0.6
# Likelihood
l1 <- p1 * p2
l1

```
The likelyhood of getting one head and one tail if the probability for 
heads is 0.5 and the probability for tails is 0.5 is `r format(l1)`

Lets look over a grid of values to find a better probability for heads.
First set up a grid of probabilities
```{r}
head_prob <-seq(0,1,0.01)
tail_prob <- 1 - head_prob
tail_prob
```
Now check the likelihood at one point on the grid.  
```{r}
#count the number of heads and the number of tails: one head , one tail
x <-c(1,1)
x[1] * head_prob[1]
x[2] * tail_prob[1]
x[1] * head_prob[1] * x[2] * tail_prob[1]
```
Create a function to do an elementary calculation, so more values can be tested.
Notice the individual probabilities are multiplied together.
```{r}
like <- function(x,p_selection){
   (head_prob[p_selection])^x[1] * tail_prob[p_selection]^x[2] 
}
like(c(1,1),1)
like(c(1,1),2)
like(c(1,1),3)
```
Use functional programming to make testing automatic
```{r}
lapply(1:100,like,x=c(1,1))
```

Consider a different data set with two head but no tails.
Try to find a larger likelihood. This is impossible

Data | Count    | Prob  | Encoding
-----|----------|-------|---------
Heads| 2        |  ?    | 2 
Tails| 0        |  ?    | 0

Lets guess that the probability of tails is close to zero.  Notice that tails never occurs
```{r}
# Encode the data with counts of heads and counts of tails
x <- c( 2,0)
lapply(1:100,like,x=x)
```


Try a different data set.

Data | Count    | Prob  | Encoding
-----|----------|-------|---------
Heads| 10        |  ?    | 10
Tails| 20        |  ?    | 20

There are twice as many tails as heads, so guess that the probability to tails is twice the probability of heads.
```{r}
x <- c( 10,20)
lapply(1:100,like,x=x)
```

Here is the maximum likelihood probabilities for heads and tails.
```{r}
head_prob[34]
tail_prob[34]
```
Exercise:
Replicate this for a three sided dice, then a six sided dice, then for the normal distribution
See ?pnorm.





