---
title: "Homework #1: Data Manipulation & Analysis using R"
output: html_notebook
---
## First load needed libraries
```{r,echo=FALSE,results='hide'}
library(Hmisc)
library(data.table)
library(tidyverse)
```

## This data is about a particular area with variables 

1. YEAR
: Year 1947 - 1962

2. PGNP
: GNP Deflator 1954=100

3. GNP
: Gross National Product in $1,000,000

4. UEM
: Unemployed in thousands

5. AF 
: Armed Forces in thousands

6. POP 
: Population in thousands

7. EM 
: Employed in thousands

Exercises:

1. Download this file from the internet link http://web.pdx.edu/~crkl/ceR/data/longley.txt and show the file as an object in R. Examine the link before you get this into R. (do not forget to use header = T, nrows=16 in the end of the link). You may want to call this data as “data”.

```{r}
d<-readr::read_table("http://web.pdx.edu/~crkl/ceR/data/longley.txt",n_max=16)
```
Assign variable labels.
```{r,eval=FALSE}
names(d)
var.labels <- c(
  YEAR = "Year 1947 - 1962", 
  PGNP = "GNP Deflator 1954=100", 
  GNP  = "Gross National Product in $1,000,000", 
  UEM  = "Unemployed in thousands", 
  AF   = "Armed Forces in thousands", 
  POP  = "Population in thousands", 
  EM   = "Employed in thousands"
)
Hmisc::upData(d,labels=var.labels)
```
3. Create a new dataset “data1” by Deleting or Replacing the variable “YEAR” with “STATES” and list them as NJ for the first 5 rows, NY for second 5 rows and PA for third 6 rows.

```{r}
STATES <- unlist(purrr::map2(c("NJ","NY","PA"),c(5,5,6),rep))
print(STATES)
d$STATES <- STATES
head(d)
```
4. Add another variable “CITY” with a name of a place that you are familiar with e.g., Edison with each row having a different city name.
```{r}
NJ<-c('Jersey City','Paterson','Trenton','Newark','Edison')
NY<-c('New York','Brooklyn','Manhattan','Albany','Queens')
PA<-c('Pennsylvania','Philadelphia','Pittsburgh','Allentown','Bethlehem','Easton')
d['CITY'] <- c(NJ,NY,PA)
head(d)
```

5. What is the minimum number of employed and average number of employed overall?
```{r}
sum_EM <- d %>% summarise(min = min(EM),mean =  mean(EM), max = max(EM))
sum_EM   
```

The minimum number unemployed is `r format(sum_EM[['min']])` thousand for any one city.

6. Which STATES has minimum unemployed and maximum unemployed?
```{r}
names(d)
EM_min <- d %>% filter(.data$EM == sum_EM[['min']] ) %>%
  select(EM,STATES)
EM_max <- d %>% filter(.data$EM == sum_EM[['max']] ) %>%
  select(EM,STATES)
print(EM_min)
print(EM_max)
```
The STATES with the minimum unemployment is `r EM_min[['STATES']]`.
The STATES with the maximum unemploymed is `r EM_max[['STATES']]`.

7. Which CITY has min GNP and the max GNP. What else do you observe? Which city you suggest if someone is planning to relocate for good? 
```{r}
sum_GNP <- d %>% summarise(min = min(GNP),mean =  mean(GNP), max = max(GNP))
d7<-d %>% filter(.data$GNP == sum_GNP[['min']] |  .data$GNP == sum_GNP[['max']])
print(d7)
```
I would relocate to the city ```r d7$CITY[2]```

8. Create a box plot by plotting AF on State. What is your understanding from the box plot. You will need to give your understanding from the box plot clearly.

```{r}
d %>% ggplot() +
  aes(x=STATES,AF) + 
  geom_boxplot()
```



9. Create a new dataset “data1” which has the following variables STATE, UEM, EM, POP.
```{r}
data1 <- d %>% select(STATES,UEM,EM,POP)
head(data1)
```
10. Create and add a new variable “diff” which is the difference of POP and EM.
```{r}
data1 <- data1 %>% mutate(diff = POP - EM)
head(data1,n = 7)
```
11. Show using scripts as to what you see in the row 7 and column 5.

```{r}
data1[c(7),c(5)]
```
12. Create a pie chart based on all the three states for UEM and comment about it.

```{r}
bp<- ggplot(data1, aes(x="", y=UEM, fill=STATES))+
geom_bar(width = 1, stat = "identity")
bp
pie <- bp + coord_polar("y", start=0)
pie
```

13. Create and show histograms for UN, UEM and POP and comment about your understanding. 
```{r}
names(d)
d %>% ggplot() + 
  aes(x=UEM) + 
  geom_bar(width = 1)
```
```{r}
d %>% ggplot() +
  aes(x=POP) + 
  geom_density()
```






